Very simplified likelihood

### 2024-08-21 inst/very_simplified_likelihood.Rmd
further work on issue #140

$$
\log\left(
(B-1)(1-p)^{B-1/2} \left[1+
\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
$$

$$
\log
(B-1) + B\log(1-p) + 
\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
$$

@@

Placing last is finishing  after $n=r$ clones finish:

\[
n=1\longrightarrow\mathcal{L}=b\]

\[
n=2\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
=\frac{1}{B(B+1)}
\]

\[
n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
\]

\[
n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{b^4}{(1+3b)(1+2b)(1+b)}=
\frac{1}{B(B+1)(B+2)(B+3)}
\]

@@

\[
\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}
\]

@@

\[
\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
\frac{(B-1)!}{(B+n-1)!}
\]

@@

```{r,label=definefsingle}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
	B <- 1/(1-a)
	if(log){
		out <- log(B-1) - lgamma(B+n) + lgamma(B+n-r-1)
	} else {  # not-log
		out <- (B-1)/prod(B+(n-r-1):(n-1))
	}
   return(out)
}

f_vec_a <- function(a,r,n, log=FALSE){  # vectorised in 'a' but not in 'r'
	sapply(a,function(a){f_single(a,r=r,n=n,log=log)})
}

f_vec_arn <- function(a,r,n,log=FALSE){
	## vectorized in 'a'; treats 'r' and 'n' as
	## vectors of independent observations

	M <- cbind(r,n)

	if(log){
		out <- 0
		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2],log=TRUE)}
	} else {
		out <- 1
		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,r=M[i,1],n=M[i,2],log=FALSE)}
	}
	return(out)
}

fapprox <- function(a,r,n,log=FALSE){
	if(log){
		return(log(B-1)-B*(r+1)/n)
	} else {
		return(B-1 - (r+1)*exp(B)/n)
	}
}
```

```{r label=defdiff}
fdash <- function(a,r,n,log=TRUE){
	B <- 1/(1-a)
	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-r-1))*B^2
	if(!log){out <- out * f_vec_arn(a,r,n,log=FALSE)}
	return(out)
}
```

@@

```{r label=verifydiff}
a <- 0.34
d <- 1e-3
r <- 4
n <- 10
c(
  numerical  = (f_single(a+d/2,r,n,log=TRUE)-f_single(a-d/2,r,n,log=TRUE))/d,
  analytical = fdash(a,r,n)
)

c(
  numerical  = (f_single(a+d/2,r,n,log=FALSE)-f_single(a-d/2,r,n,log=FALSE))/d,
  analytical = fdash(a,r,n,log=FALSE)
)
```

@@

```{r,label=defmle}
MLE <- function(r,n,give=FALSE){
	d <- 1e-6
	out <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
	if(!give){out <- out$maximum}
	return(out)
}

MLEa <- function(r,n,give=FALSE,small=1e-4){
	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(small,1-small))
	if(!give){out <- out$root}
	return(out)
}

@@

rbind(
showdiff(MLE(1,9),MLEa(1,9)),
showdiff(MLE(3,9),MLEa(3,9)),
showdiff(MLE(3,7),MLEa(3,7))
)
```

@@

```{r,label=verifytwo}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
r <- 3
c(
way1 = b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
way2 = (B-1)/prod(B+(5:8)),
way3 = f_vec_arn(a,r,n,log=FALSE)
)
```

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(r in 0:n){
    if(r==0){
      jj <- 1
    } else if(r==n){n
      jj <- 0
    } else {
      jj <- MLE(r,n)
    }
    out[[n]][r+1] <- jj
  }
}
out
```

```{r,label=plotjj}
plotterp <- function(...){
plot(NA,xlim=c(1,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
for(n in 1:10){
  for(r in 0:n){   points(n,out[[n]][r+1],pch=16) }
  for(r in 0:n){   text(n+0.2,out[[n]][r+1],r,cex=0.5,col='gray') }
}
}
plotterp()
pdf(file="dotprobs.pdf")
plotterp()
dev.off()
```

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 8
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
grid()
rain <- rainbow(n+1)
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rain[r+1])
}
abline(v=MLE(r=4,n=8))
text(0.07,0.95,"r=8",col=rain[9])
text(0.20,0.95,"r=7",col=rain[8])
text(0.26,0.91,"r=6",col=rain[7])
text(0.29,0.83,"r=5",col=rain[6])
text(0.33,0.76,"r=4",col=rain[5])
text(0.35,0.64,"r=3",col=rain[4])
text(0.39,0.53,"r=2",col=rain[3])
text(0.45,0.41,"r=1",col=rain[2])
text(0.63,0.22,"r=0",col=rain[1])
}

@@

```{r shown4n8mle}
MLE(r=4,n=8)
```

@@

```{r checofirst}
checo_like <- function(a){
  out <- a*0
  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year)-1,getnum(year)-1,log=TRUE) }
  return(out)
}

@@

```{r definef1log}
f1_logistic <- function(vec){
  alpha <- vec[1]
  beta  <- vec[2]
  out <- 0
  for(year in 2011:2023){
     LO <- alpha + beta*(year-2011)
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,getfoc(year)-1,getnum(year)-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r findcredint}
(a_lower <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.3,0.5))$root)
(a_upper <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.5,0.6))$root)
```

@@

```{r calculatelikes1,cache=F}
a <- seq(from=0.005,to=0.85,by=0.01)
La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b',ylab='fishy')
abline(h=c(0,-2))
```

```{r splitresults}
supp2 <- function(vec,place,runners){
  a_stirling <- vec[1]
  a_auckland <- vec[2]
  out <- 0
  for(i in 1:3){
    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
  }
  for(i in 4:10){
    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
```

```{r definesupp2ed}
supp2_ed <- function(vec,place1,place2,runners1,runners2){
  out <- 0
  M1 <- cbind(place1,runners1)
  M2 <- cbind(place2,runners2)
  for(i in seq_len(nrow(M1))){
    out <- out + f_vec_arn(vec[1],r=M1[i,1]-1,n=M1[i,2]-1,log=TRUE)
  }
  for(i in seq_len(nrow(M2))){
    out <- out + f_vec_arn(vec[2],r=M2[i,1]-1,n=M2[i,2]-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r,label=defsupp2}
jj1 <-
optim(
   c(.4,.8),
   function(vec){
      supp2_ed(vec,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
   control=list(fnscale = -1)
)
jj1
```

>>>>

$$
\log\left(
(B-1)(1-p)^{B-1/2} \left[1+
\frac{(1-6B+6B^2)p}{12(1-p)n}\right]\right)
$$

$$
\log
(B-1) + B\log(1-p) +
n^{-1}
\frac{(1-6B+6B^2)p}{12(1-p)} +{\mathcal O}\left(n^{-2}\right)
$$

$$
\log
(B-1) + B\log(1-p) +
\frac{1}{12}n^{-1}\frac{p}{1-p}(1-6B+6B^2) + {\mathcal O}\left(n^{-2}\right)
$$

@@

Placing last is finishing after $n=r$ clones finish, that is, $R=n+1$
[me plus two clones, I come third].

\[
n=1,R=2\longrightarrow\mathcal{L}=b\]

\[
n=2,R=3\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
=\frac{1}{B(B+1)}
\]

\[
n=3,R=4\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
\]

\[
n=4,R=n+1\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{b^4}{(1+3b)(1+2b)(1+b)}=
\frac{1}{B(B+1)(B+2)(B+3)}
\]

@@

\[
\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+2)}
\]

@@

\[
\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
\frac{(B-1)!}{(B+n-1)!}=
\frac{\Gamma(B)}{\Gamma(B+n)}
\]

@@

```{r,label=definefsingle}
f_single <- function(a,R,n,log=FALSE){  # not vectorised
    stopifnot(R>0)
	B <- 1/(1-a)
	if(log){
		out <- log(B-1) + lgamma(B+n-R) - lgamma(B+n)
	} else {  # not-log
		out <- (B-1)/prod(B+(n-R):(n-1))
	}
   return(out)
}

f_vec_a <- function(a,R,n, log=FALSE){  # vectorised in 'a' but not in 'R'
	sapply(a,function(a){f_single(a,R=R,n=n,log=log)})
}

f_vec_arn <- function(a,R,n,log=FALSE){
	## vectorized in 'a'; treats 'R' and 'n' as
	## vectors of independent observations

	M <- cbind(R,n)

	if(log){
		out <- 0
		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,R=M[i,1],n=M[i,2],log=TRUE)}
	} else {
		out <- 1
		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,R=M[i,1],n=M[i,2],log=FALSE)}
	}
	return(out)
}

fapprox <- function(a,R,n,log=FALSE){
	if(log){
		return(log(B-1)-B*R/n)
	} else {
		return((B-1)/exp(R*B/n))
	}
}
```

```{r label=defdiff}
fdash <- function(a,R,n,log=TRUE){
	B <- 1/(1-a)
	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-R))*B^2
	if(!log){out <- out * f_vec_arn(a,R,n,log=FALSE)}
	return(out)
}
```

@@

```{r label=verifydiff}
a <- 0.34
d <- 1e-3
R <- 4
n <- 10
c(
  numerical  = (f_single(a+d/2,R,n,log=TRUE)-f_single(a-d/2,R,n,log=TRUE))/d,
  analytical = fdash(a,R,n)
)

c(
  numerical  = (f_single(a+d/2,R,n,log=FALSE)-f_single(a-d/2,R,n,log=FALSE))/d,
  analytical = fdash(a,R,n,log=FALSE)
)
```

@@

```{r,label=defmle}
MLE <- function(R,n,give=FALSE){
	d <- 1e-6
	out <- optimize(f_vec_arn,c(d,1-d),R=R,n=n,log=TRUE,maximum=TRUE)
	if(!give){out <- out$maximum}
	return(out)
}

MLEa <- function(R,n,give=FALSE,small=1e-4){
	out <- uniroot(f=function(a){fdash(a,R,n)},interval=c(small,1-small))
	if(!give){out <- out$root}
	return(out)
}

@@

rbind(
showdiff(MLE(3,9),MLEa(3,9)),
showdiff(MLE(3,7),MLEa(3,7))
)
```

@@

```{r,label=verifytwo}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
R <- 4
c(
way1 = b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
way2 = (B-1)/prod(B+(5:8)),
way3 = f_vec_arn(a,R,n,log=FALSE)
)
```

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(R in seq_len(n+1)){
    if(R==1){
      jj <- 1
    } else if(R == n+1){
      jj <- 0
    } else {
      jj <- MLE(R,n)
    }
    out[[n]][R] <- jj
  }
}
out
```

```{r,label=plotjj}
plotterp <- function(...){
plot(NA,xlim=c(1,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
for(n in 1:10){
  for(R in seq_len(n+1)){   points(n,out[[n]][R],pch=16) }
  for(R in seq_len(n+1)){   text(n+0.2,out[[n]][R],R,cex=0.5,col='gray') }
}
}
plotterp()
pdf(file="dotprobs.pdf")
plotterp()
dev.off()
```

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 8
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
grid()
rain <- rainbow(n+1)
for(R in 1:n){
  y <- f_vec_a(a,R,n,log=FALSE)
  if(R==1){y[length(y)] <- 1}
  y <- y/max(y,na.rm=TRUE)
  if(R>1){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rain[R])
}
abline(v=MLE(R=5,n=8))
text(0.07,0.95,"R=9",col=rain[9])
text(0.07,0.95,"R=8",col=rain[8])
text(0.20,0.95,"R=7",col=rain[7])
text(0.26,0.91,"R=6",col=rain[6])
text(0.29,0.83,"R=5",col=rain[5])
text(0.33,0.76,"R=4",col=rain[4])
text(0.35,0.64,"R=3",col=rain[3])
text(0.39,0.53,"R=2",col=rain[2])
text(0.45,0.41,"R=1",col=rain[1])
}

@@

```{r shown4n8mle}
MLE(R=5,n=8)
```

@@

```{r checofirst}
checo_like <- function(a){
  out <- a*0
  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)-1,log=TRUE) }
  return(out)
}

@@

```{r definef1log}
f1_logistic <- function(vec){
  alpha <- vec[1]
  beta  <- vec[2]
  out <- 0
  for(year in 2011:2023){
     LO <- alpha + beta*(year-2011)
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,getfoc(year),getnum(year)-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r findcredint}
(a_lower <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.1,0.5))$root)
(a_upper <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.5,0.9))$root)
```

@@

```{r calculatelikes1,cache=F}
a <- seq(from=0.005,to=0.85,by=0.01)
La <- f_vec_arn(a,R=results$place[1],results$runners[1]-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b',ylab='fishy')
abline(h=c(0,-2))
```

```{r splitresults}
supp2 <- function(vec,place,runners){
  a_stirling <- vec[1]
  a_auckland <- vec[2]
  out <- 0
  for(i in 1:3){
    out <- out + f_vec_a(a_stirling,R=place[i],runners[i]-1,log=TRUE)
  }
  for(i in 4:10){
    out <- out + f_vec_a(a_auckland,R=place[i],runners[i]-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
```

```{r definesupp2ed}
supp2_ed <- function(vec,place1,place2,runners1,runners2){
  out <- 0
  M1 <- cbind(place1,runners1)
  M2 <- cbind(place2,runners2)
  for(i in seq_len(nrow(M1))){
    out <- out + f_vec_arn(vec[1],R=M1[i,1],n=M1[i,2]-1,log=TRUE)
  }
  for(i in seq_len(nrow(M2))){
    out <- out + f_vec_arn(vec[2],R=M2[i,1],n=M2[i,2]-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r,label=defsupp2}
jj1 <-
optim(
   c(.4,.8),
   function(vec){
      supp2_ed(vec,rank[wp],rank[wa],class_size[wp]-1,class_size[wa]-1)},
   control=list(fnscale = -1)
)
jj1
```
